---
title: "PrEP Clusters"
author: "Randy Yee (pcx5@cdc.gov)"
date: "2025-01-15"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(janitor)
library(factoextra)
library(cluster)
library(corrplot)

knitr::opts_chunk$set(echo = TRUE)
options(scipen = 999)
```

# Import and Assemble Subsets

## Import the PSNUxIM dataset

```{r}
setwd("~/PrEP Country Clusters")
Genie_PSNU_IM_Global_frozen_2025_01_13 <- read_delim("Genie-OU_IM-Global-frozen-2025-01-14.txt", 
                                                     delim = "\t", escape_double = FALSE, 
                                                     trim_ws = TRUE)
```

## Add a pop column then group by to remove all the unnecessary columns

```{r}
df <- Genie_PSNU_IM_Global_frozen_2025_01_13 %>%
  filter(typemilitary == "N") %>%
  mutate(population = case_when(standardizeddisaggregate == "Total Numerator" ~ "Total",
                                is.na(ageasentered) & is.na(sex) ~otherdisaggregate_sub,
                                TRUE ~paste0(ageasentered, "-", sex)))%>%
  group_by(country, 
           #snu1, snu1uid, 
           fiscal_year,
           indicator, standardizeddisaggregate, otherdisaggregate, otherdisaggregate_sub, ageasentered, population, sex) %>%
  summarize(cumulative = sum(cumulative, na.rm =T),
            target = sum(targets, na.rm = T)) %>%
  ungroup()
```

## Achievement Subset
 
```{r}
df_achieve <- df %>%
  filter(fiscal_year == 2024) %>%
  #filter(standardizeddisaggregate!="Total Numerator") %>%
  filter(target != 0) %>%
  group_by(country, 
           #snu1, snu1uid, 
           indicator, population,
           fiscal_year) %>%
  summarize(cumulative = sum(cumulative, na.rm =T),
            target = sum(target, na.rm = T)) %>%
  ungroup() %>%
  mutate(achievement = cumulative/target) %>%
  mutate(indicator = str_replace_all(indicator,"_","")) %>%
  mutate(label = paste0(indicator, " Achievement", "_", population)) %>%
  filter(indicator != "HTSTSTNEG") %>%
  select(-c(cumulative, target, indicator, population, fiscal_year)) %>%
  pivot_wider(names_from = label, values_from = achievement) %>%
  filter(!is.na(`PrEPNEW Achievement_Total`))
```
 
## PrEP-to-Need and Coverage Subset
 
```{r}
df_ptn <- df %>%
  filter(fiscal_year == 2024) %>%
  filter(standardizeddisaggregate%in% c("Total Numerator",
                                        "Modality/Age/Sex/Result", "Modality/Age Aggregated/Sex/Result", "KeyPop/Result",
                                        "Age/Sex", "KeyPop", "KeyPopAbr")) %>%
  select(country, 
         #snu1, snu1uid, 
         indicator, population, cumulative) %>%
  pivot_wider(names_from = indicator, values_from = cumulative, values_fn = sum) %>%
  mutate(PTN = (PrEP_NEW+PrEP_CT)/HTS_TST_POS,
         `PrEP Coverage` = PrEP_NEW/HTS_TST_NEG) %>%
  select(-c(PrEP_NEW, PrEP_CT, HTS_TST_POS, HTS_TST_NEG)) %>%
  pivot_longer(PTN:`PrEP Coverage`, names_to = "label", values_to = "value") %>%
  filter(!is.na(value) | value != Inf) %>%
  mutate(label = paste0(label, "_", population)) %>%
  select(-population) %>%
  pivot_wider(names_from = label, values_from = value)
```
 
## PrEP Type Subset

```{r}
df_preptype <- df %>%
  filter(fiscal_year == 2024) %>%
  filter(indicator %in% c("PrEP_NEW", "PrEP_CT")) %>%
  filter(standardizeddisaggregate=="Type") %>%
  select(-target) %>%
  pivot_wider(names_from = otherdisaggregate, values_from = cumulative, values_fn = sum) %>%
  mutate(Total = rowSums(.[9:13], na.rm = T),
        `Oral Type` =  `N/A, PrEP Type - Oral`/Total,
        `Oral Type1` =  `PrEP Type - Oral`/Total,
        `Other Type` = `PrEP Type - Other`/Total,
        `Injectable Type` = `N/A, PrEP Type - Injectable`/Total,
        `Injectable Type1` = `PrEP Type - Injectable`/Total,
        ) %>%
  pivot_longer(Total:`Injectable Type1`, names_to = "label", values_to = "value") %>%
  filter(label != "Total") %>%
  select(country, indicator, label, value) %>%
  mutate(label = case_when(label == "Oral Type1" ~ "Oral Type",
                           label == "Injectable Type1" ~ "Injectable Type",
                           TRUE ~label)) %>%
  mutate(indicator = str_replace_all(indicator,"_","")) %>%
  unite("label", c(indicator, label), sep = " ") %>%
  mutate(label = paste0(label, "_Total")) %>%
  pivot_wider(names_from = label, values_from = value, values_fn = function(x) sum(x, na.rm = TRUE))
```

## PrEP_CT Result Subset

```{r}
df_ctresult <- df %>%
  filter(fiscal_year == 2024) %>%
  filter(standardizeddisaggregate == "TestResult") %>%
  select(-target) %>%
  pivot_wider(names_from = otherdisaggregate, values_from = cumulative) %>%
  mutate(`PrEPCT Percent Positive_Total` = `PrEP Test Result - Positive`/sum(`PrEP Test Result - Positive`, `PrEP Test Result - Negative`, `PrEP Test Result - Other`, na.rm = T)) %>%
  select(country, `PrEPCT Percent Positive_Total`)
```

## KP Proportion Subset

```{r}
df_kp <- df %>%
  filter(fiscal_year == 2024) %>%
  filter(indicator %in% c("PrEP_NEW", "PrEP_CT")) %>%
  filter(population %in% c("Total", "MSM", "FSW", "TG", "PWID", "People in prisons and other enclosed settings")) %>%
  group_by(country, indicator, population) %>%
  summarize(cumulative = sum(cumulative, na.rm = T)) %>%
  ungroup() %>%
  pivot_wider(names_from = population, values_from = cumulative) %>%
  mutate(`Percent_MSM` = MSM/Total,
         `Percent_FSW` = FSW/Total,
         `Percent_TG` = TG/Total,
         `Percent_PWID` = PWID/Total,
         `Percent_People in prisons and other enclosed settings` = `People in prisons and other enclosed settings`/Total) %>%
  select(country, indicator, `Percent_MSM`:`Percent_People in prisons and other enclosed settings`) %>%
  pivot_longer(`Percent_MSM`:`Percent_People in prisons and other enclosed settings`, names_to = "label", values_to = "value") %>%
  mutate(indicator = str_replace_all(indicator,"_","")) %>%
  unite("label", c(indicator, label), sep = " ") %>%
  pivot_wider(names_from = label, values_from = value)

df_kp[is.na(df_kp)] <- 0
  
```

## HIV Positive Change Subset

```{r}
df_pos <- df %>%
  filter(indicator == "HTS_TST_POS") %>%
  select(country, fiscal_year, population, cumulative) %>%
  pivot_wider(names_from = fiscal_year, values_from = cumulative, values_fn = function(x) sum(x, na.rm = TRUE)) %>% 
  mutate(`Positive Change` = (`2024`-`2023`)/`2023`) %>%
  # filter(!is.na(`Positive Change`)) %>%
  mutate(population = paste0("Positive Change_", population)) %>%
  select(-c(`2024`, `2023`)) %>%
  pivot_wider(names_from = population, values_from = `Positive Change`)
  #filter(!is.nan(Per_Pos) | !is.na(Per_Pos) | Per_Pos != Inf)
```

## Testing Yield Subset

```{r}
df_yield <- df %>%
  filter(fiscal_year == 2024) %>%
  filter(indicator %in% c("HTS_TST_POS", "HTS_TST_NEG")) %>%
  select(country, indicator, population, cumulative) %>%
  pivot_wider(names_from = indicator, values_from = cumulative, values_fn = function(x) sum(x, na.rm = TRUE)) %>%
  mutate(Yield = HTS_TST_POS/(HTS_TST_POS + HTS_TST_NEG)) %>%
  mutate(population = paste0("Yield_", population)) %>%
  select(-c(HTS_TST_POS, HTS_TST_NEG)) %>%
  pivot_wider(names_from = population, values_from = Yield)
  #filter(!is.nan(Yield) | !is.na(Yield) | Yield != Inf)
```

TODO: Gini?

## Merge Subsets

```{r}
df1 <- df_ptn %>%
  left_join(df_achieve) %>%
  left_join(df_preptype) %>%
  left_join(df_ctresult) %>%
  left_join(df_kp) %>%
  left_join(df_pos) %>%
  left_join(df_yield)
```

Remove any empty rows and columns

```{r}
df2 <- df1

df2[df2 == 0] <- NA

df2 <- df2 %>%
  janitor::remove_empty("cols", cutoff = 0.9) # Higher
```

# Decision Tree 

```{r}
library(rpart)
library(rpart.plot)
library(caTools)
```

```{r}
df_rf <- df1 %>%
  pivot_longer(2:200, names_to = "label", values_to = "value") %>%
  separate(label, into = c("indicator", "population"), sep = "_") %>%
  pivot_wider(names_from = indicator, values_from = value)
```

```{r}
df_rf <- df2 %>%
  select(-c(50:65, 67:85))
```

```{r}
set.seed(42)
sample_split <- sample.split(Y = df_rf$`Positive Change_Total`, SplitRatio = 0.75)
train_set <- subset(x = df_rf, sample_split == TRUE)
test_set <- subset(x = df_rf, sample_split == FALSE)
```

```{r}
model <- rpart(`Positive Change_Total` ~ ., data = train_set, method = "class")
model
```

```{r}
rpart.plot(model)
```

```{r}
importances <- varImp(model)
importances %>%
  arrange(desc(Overall))
```




# Cluster

```{r}
test <- df2[complete.cases(df2)==T,]

df_m <- test %>%
  column_to_rownames("country")
```

## Correlation

```{r}
corr <- cor(df_m)
corr
```

```{r}
corr_simple <- function(data=df,sig=0.5){
  #convert data to numeric in order to run correlations
  #convert to factor first to keep the integrity of the data - each value will become a number rather than turn into NA
  df_cor <- data %>% mutate_if(is.character, as.factor)
  df_cor <- df_cor %>% mutate_if(is.factor, as.numeric)
  #run a correlation and drop the insignificant ones
  corr <- cor(df_cor)
  #prepare to drop duplicates and correlations of 1     
  corr[lower.tri(corr,diag=TRUE)] <- NA 
  #drop perfect correlations
  corr[corr == 1] <- NA 
  #turn into a 3-column table
  corr <- as.data.frame(as.table(corr))
  #remove the NA values from above 
  corr <- na.omit(corr) 
  #select significant values  
  corr <- subset(corr, abs(Freq) > sig) 
  #sort by highest correlation
  corr <- corr[order(-abs(corr$Freq)),] 
  #print table
  print(corr)
  #turn corr back into matrix in order to plot with corrplot
  mtx_corr <- reshape2::acast(corr, Var1~Var2, value.var="Freq")
  
  #plot correlations visually
  corrplot(mtx_corr, is.corr=FALSE, tl.col="black", na.label=" ")
}
corr_simple(df_m)

corrplot(corr, tl.col = "brown", tl.srt = 30, bg = "White",
         title = "\n\n Correlation Plot Of KPIs",
         type = "upper")
```


## Optimal # of Clusters

```{r}
distance <- get_dist(df_m)
fviz_dist(distance, gradient = list(low = "#00AFBB", mid = "white", high = "#FC4E07"))
```
```{r}
set.seed(123)

fviz_nbclust(df_m, kmeans, method = "wss")
```


```{r}
fviz_nbclust(df_m, kmeans, method = "silhouette")
```


```{r}
gap_stat <- cluster::clusGap(df_m, FUN = kmeans, nstart = 25,
                    K.max = 10, B = 50)
fviz_gap_stat(gap_stat)
```
```{r}
res <- NbClust::NbClust(df_m, distance = "euclidean", min.nc=2, max.nc=10, 
            method = "kmeans", index = "ch")
            
res$All.index
res$Best.nc
res$Best.partition
```
```{r}
fviz_nbclust(res)
```
## K-Means Cluster

```{r}
clusters <- kmeans(df_m, 4)
```

```{r}
fviz_cluster(clusters, data = df_m) +
  theme_minimal()
```
```{r}
results <- df_m %>%
  mutate(Cluster = clusters$cluster)

results_s <- df_m %>%
  mutate(Cluster = clusters$cluster) %>%
  group_by(Cluster) %>%
  summarise_all("median")
```

```{r}
write.csv(results, "results.csv")
write.csv(results_s, "summarizedresults.csv")
```

## Hierarchical Cluster

```{r}
hresults <- hclust(dist(df_m))
plot(hresults)
```
```{r}
cutree(hresults, 6)
```

```{r}
hresults2 <- hclust(dist(df_m), method = 'average')
plot(hresults2)
```

```{r}
m <- c( "average", "single", "complete", "ward")
names(m) <- c( "average", "single", "complete", "ward")

# function to compute coefficient
ac <- function(x) {
  agnes(df_m, method = x)$ac
}

map_dbl(m, ac)
```
```{r}
hc3 <- agnes(df_m, method = "ward")
pltree(hc3, cex = 0.6, hang = -1, main = "Dendrogram of agnes") 
```
```{r}
# Ward's method
hc5 <- hclust(dist(df_m), method = "ward.D2" )

# Cut tree into 4 groups
sub_grp <- cutree(hc5, k = 6)

# Number of members in each cluster
table(sub_grp)
```

```{r}
df_m %>%
  mutate(cluster = sub_grp)
```

